---
title: "Mixed ANOVA"
output: learnr::tutorial
runtime: shiny_prerendered
description: >
  Mixed ANOVA
---

```{r setup, include=FALSE}
library(learnr)
library(knitr)
library(lsr)
library(car)
library(ggplot2)
library(ANOVATutorials)
library(dplyr)
library(ez)
library(lme4)
```

# Mixed ANOVA

## Overview

This tutorial focuses on mixed ANOVA - ANOVA with one or more within subject factors and one or more between subject factors. I also discuss effect sizes and simple interaction tests.

NOTE: THIS TUTORIAL RENDERS BEST IN A BROWSER WINDOW. Use the button in the upper left hand corner to open in browser window. 

The PowerPoint slides for the presentation in the videos are on Canvas if you want a standalone copy. 

The videos (as well as others) can also be found on my YouTube channel 
 https://www.youtube.com/channel/UC5kDZTyHZlgSgSEa3YQXOi

### Learning Objectives

Expand knowledge regarding withing subjects ANOVA to designs that add a between subject factor

Explain the conditions for applying mixed ANOVA

Compute effect sizes from linear mixed models

Apply the above procedures to making decisions about null hypotheses

Graph and interpret interaction effects 

Use R to compute ANOVA and eta-squared

Apply simple interactions tests to formally evaluate three-way interactions


## Data

*West* is the datafile used in exercises. These data examine changes in body image following an intervention from **pre** to **post**. Half of the participants interacted in a bar setting with other participants (the bar was otherwise empty) while naked and the others interacted similarly (in a different room) while clothed **condition**.

*Gaither* examines support for Hillary Clinton before (**pre**) and after (**post**) the 2016 election and how such relationships are affected by participant gender (**gender**).

Both datafiles will be used for several examples. 

### Packages 

This tutorial uses the following new packages:

* `lsr` for effect sizes
* `car` for ANOVA computations
* `ez` for running ANOVA
* `lme4` for linear mixed models

## Video I Introduction

![](https://youtu.be/4i0znRXwqJ8)

## Quiz 1 

```{r echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
``` 

```{r quiz1, ECHO=F}
learnr::quiz(
  learnr::question("Mixed ANOVA involves ... ",
    learnr::answer("Two measurement points"),
    learnr::answer("At least one between and one within subject factor", correct = TRUE),
    learnr::answer("Mixing stuff up"),
    learnr::answer("A standup mixer"),
    correct = "Correct!",
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  ),
  learnr::question("Sphericity considerations now include ...",
    learnr::answer("Any factor or interaction that contains at least one within subjects term", correct = TRUE),
    learnr::answer("All of the factors"),
    learnr::answer("None of the factors"),
    learnr::answer("Mixed ANOVA does not assume sphericity"),
    correct = "Correct.",
    random_answer_order = TRUE,
    incorrect = "Sorry, that is incorrect. Try again.",
    allow_retry = T
  )
)

``` 

## Video II Mixed ANOVA

![](https://youtu.be/flGtnxym2E0)  

## Quiz

```{r quiz2, ECHO=F}
learnr::quiz(
 learnr::question("Simple interactions tests ... ",
    learnr::answer("Break down complex interactions between three or more factors", correct = TRUE),
    learnr::answer("Break down complex interactions between two factors"),
    learnr::answer("Test main effects"),
    learnr::answer("Breakdown SS within"),  
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  ),
   learnr::question("How would you set up a simple interactions test for three factors?",
    learnr::answer("Do t-tests on all pairs"),
    learnr::answer("Using ANOVA"),
    learnr::answer("with the computer"),
    learnr::answer("Examine the interactions between two factors at each of the levels of the other factor", correct = TRUE),  
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
)
)

``` 


## Exercise 

```{r echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Mixed subjects ANOVA requires the same data restructuring as for within. For most applications, we use "wide" format wherein every row represents one person and we have a column for each variable. Within subjects approaches generally require "long" format where each person will be in multiple rows (one for each level of the within subject factor).

The first line of code creates an ID number for us. This is essential because it is the only way we know who goes with what. 

The second moves us from wide to long format. key is the name for a new variable that establishes the within subjects condition and value is what we are calling our new DV. -ID makes sure ID is repeated each time a person had data. 

Next we need to make sure all factors are converted to factors and all numeric data are converted to numeric. 

The final line runs the anova using the `ez` package. dv, wid, and within are the relevant variables for analysis.  

```{r}
Gaither$ID<-seq(1:191)
Gaitherlong<-tidyr::gather(Gaither,key="Condition",value="Censor",-ID, -gender)
Gaitherlong$Censor<-as.numeric(Gaitherlong$Censor)
Gaitherlong$Condition<-as.factor(Gaitherlong$Condition)
modelGaither<-ez::ezANOVA(data=Gaitherlong, dv=.(Censor), wid=.(ID), within = .(Condition),between=.(gender), type=3, detailed=TRUE)
modelGaither
```

For simple effects we simply need to do a couple of paired t-tests. This requires subsetting. 

```{r}
men<-subset(Gaither, gender=="Men")
women<-subset(Gaither, gender=="Women")
xx<-t.test(men$pre,men$post,paired=TRUE)
xx
lsr::cohensD(men$pre,men$post,method="paired")

```


### Exercise 

Using the *Westlong* data, predict attitudes body_image from **time** and **condition** (naked or not naked). The data are already restructured. 

```{r ex2, exercise = TRUE, exercise.lines = 7}

```

```{r ex2-solution}
modelWest<-ez::ezANOVA(data=Westlong, dv=.(body_image), wid=.(id), within = .(time),between=.(condition), type=3, detailed=TRUE)
modelWest
```

### Simple Effects

Since there are only two factors, we only need to perform simple effects tests, in the form of two paired t-tests, subset by condition. Use the *West* not the *Westlong* data. 

```{r ex3, exercise = TRUE, exercise.lines = 6}

```


```{r ex3-solution}
nekkid<-subset(West, condition=="Naked")
not_nekkid<-subset(West, condition=="Control")
t.test(nekkid$pre,nekkid$post,paired=TRUE)
lsr::cohensD(nekkid$pre,nekkid$post,method="paired")
t.test(not_nekkid$pre,not_nekkid$post,paired=TRUE)
lsr::cohensD(not_nekkid$pre,not_nekkid$post,method="paired")

```

## Quiz

```{r}
modelWest<-ez::ezANOVA(data=Westlong, dv=.(body_image), wid=.(id), within = .(time),between=.(condition), type=3, detailed=TRUE)
modelWest
tapply(West$pre, list(West$condition),mean)
tapply(West$post, list(West$condition),mean)
nekkid<-subset(West, condition=="Naked")
not_nekkid<-subset(West, condition=="Control")
t.test(nekkid$pre,nekkid$post,paired=TRUE)
lsr::cohensD(nekkid$pre,nekkid$post,method="paired")
t.test(not_nekkid$pre,not_nekkid$post,paired=TRUE)
lsr::cohensD(not_nekkid$pre,not_nekkid$post,method="paired")
```

```{r echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r quiz5, ECHO=F}
learnr::quiz(
 learnr::question("What is your primary conlusion regarding the interaction?",
    learnr::answer("Body image rose from pretest to posttest"),
    learnr::answer("Body image fell from pretest to posttest"),
    learnr::answer("The posttest happened after the pretest"),
    learnr::answer("Body image changed in a different manner for each group",correct=TRUE),  
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  ),
 learnr::question("What do the simple effects tests tell us about the interaction?",
    learnr::answer("Body image rose from pretest to posttest"),
    learnr::answer("Body image fell from pretest to posttest"),
    learnr::answer("The posttest happened after the pretest"),
    learnr::answer("Body image improved for the nakeds but got worse for those in the control group",correct=TRUE),  
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  )
)

``` 


## Video 3: APA style

![](https://youtu.be/spn_-2mVOgc)

##  

Congratulations! You've reached the end of the tutorial. 
