---
title: "Factorial ANOVA"
output: learnr::tutorial
runtime: shiny_prerendered
description: >
  Two Factor ANOVA, interactions, effect size, assumptions, and simple effects 
---

```{r setup, include=FALSE}
library(learnr)
library(knitr)
library(lsr)
library(car)
library(ggplot2)
library(ANOVATutorials)
library(BayesFactor)
library(phia)
library(dplyr)

```

# Factorial ANOVA

## Overview

This tutorial focuses on the two factor ANOVA with a particular emphasis on interpreting interactions. I also discuss effect sizes, simple effects tests, assumptions, Bayes Factors, and APA style.

NOTE: THIS TUTORIAL RENDERS BEST IN A BROWSER WINDOW. Use the button in the upper left hand corner to open in browser window. 

The PowerPoint slides for the presentation in the videos are on Canvas if you want a standalone copy. 

The videos (as well as others) can also be found on my YouTube channel https://www.youtube.com/channel/UC5kDZTyHZlgSgSEa3YQXOi

Credit to our oldest, Ernesto Aberson for his work creating the Jewish Space Lasers thumbnail. Credit also to GOP Congresswoman Marjorie Taylor Greene - this artwork would not exist if not for her crazy beliefs. 

### Learning Objectives

Define the different types of Sums of Square in the two factor context

Explain the conditions for applying the Two Factor ANOVA

Compute Sums of Squares, degrees of freedom, Mean Square, and F

Given a problem, define the null and alternative hypotheses

Apply the above procedures to making decisions about null hypotheses

Describe and calculate partial eta-squared as a measure of effect size

Graph and interpret interaction effects 

Use R to compute ANOVA and eta-squared

Excitedly tell your friends and family about interaction effects

Apply simple effects tests to formally evaluation interactions

Test assumptions for a factorial ANOVA

Get really worked up about why partial eta-squared is terrible

## Data

*Green1* is the datafile used in exercises. These data how goals and partner friendship network influence a number of perceptions. These data include two factors, **goal_manip** and **fr_manip** and two DVs positive emotions (**XPosEmo**) and overall attitude measure (**XMeta**).

Green, D. J., Wout, D. A., & Murphy, M. C. (in press). Learning goals mitigate identity threat for Black individuals in threatening interracial interactions. *Cultural Diversity and Ethnic Minority Psychology.* https://doi.org/10.1037/cdp000033

### Packages 

This tutorial uses the following new packages:

* `lsr` for effect sizes
* `car` for ANOVA computations
* `ggplot2` for graphing 
* `BayesFactor` for ... I shouldn't have to tell you what the `BayesFactor` does.
* `phia` for simple effects and making all the work I did writing code to calculate simple effects in previous semesters obsolete. 

## Video I Introduction

![](https://youtu.be/dP9PXMm_HvM)

## Quiz 1 

```{r echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
``` 

```{r quiz1, ECHO=F}
learnr::quiz(
  learnr::question("The most important effect in factorial ANOVA usually is ... ",
    learnr::answer("The interaction", correct = TRUE),
    learnr::answer("The main effects"),
    learnr::answer("The intercept"),
    correct = "Correct!",
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  ),
  learnr::question("What new sums squares value(s) do we see in factorial ANOVA? ",
    learnr::answer("SS A, SS B, and SS AxB", correct = TRUE),
    learnr::answer("SS Factorial"),
    learnr::answer("SS Between"),
    learnr::answer("SS Within"),
    correct = "Correct.",
    random_answer_order = TRUE,
    incorrect = "Sorry, that is incorrect. Try again.",
    allow_retry = T
  ),
  learnr::question("Which of the following describes an interaction effect? ",
    learnr::answer("The levels of A1 and A2 differ"),
    learnr::answer("There is a differential effect of A across the levels of B", correct = TRUE),
    learnr::answer("There is the same effect of A across the levels of B"),
    learnr::answer("There is no difference between B1 and B2"),    
    correct = "Correct.",
    random_answer_order = TRUE,
    incorrect = "Sorry, that is incorrect. Try again.",
    allow_retry = T
  )
)

``` 

## Video II Calculations

![](https://youtu.be/jNP2jcrdtZw)  

## Exercise Calculations

$\large SS_{Total}=\Sigma(x-\bar{x}_t)^2$  
$\large SS_{Between}=\Sigma n_{cell}(\bar{x}_{cell}-\bar{x}_t)^2$  
$\large SS_{Within}=\Sigma(x-\bar{x}_{cell})^2$    
$\large SS_{A}=\Sigma n_{A}(\bar{x}_{A}-\bar{x}_t)^2$ 
$\large SS_{B}=\Sigma n_{B}(\bar{x}_{B}-\bar{x}_t)^2$ 
$\large SS_{AxB}=SS_{BG}-SS_{A}-SS_{B}$  

$\large MS_{Between}=\frac{SS_{Between}}{df_{Between}}$  

$\large MS_{Within}=\frac{SS_{Within}}{df_{Within}}$  

$\large \eta^2=\frac{SS_{Between}}{SS_{Total}}$  

$\large MS_{A}=\frac{SS_{A}}{df_{A}}$  
$\large df_{A}=A-1$  
$\large F_{A}=\frac{MS_A}{df_A}$  
$\large MS_{B}=\frac{SS_{B}}{df_{B}}$  
$\large df_{B}=B-1$  
$\large F_{B}=\frac{MS_B}{df_B}$  
$\large MS_{AxB}=\frac{SS_{AxB}}{df_{AxB}}$  
$\large df_{AxB}=(A-1)(B-1)$  
$\large F_{AxB}=\frac{MS_{AxB}}{df_{AxB}}$  

$\large MS_{Within}=\frac{SS_{Within}}{df_{Within}}$    

$\large F = \frac{MS_{Between}}{MS_{Within}}$  

$\large \eta^2_A=\frac{SS_{A}}{SS_{Total}-SS_B-SS_{AxB}}$  
$\large \eta^2_B=\frac{SS_{B}}{SS_{Total}-SS_A-SS_{AxB}}$  
$\large \eta^2_{AxB}=\frac{SS_{AxB}}{SS_{Total}-SS_A-SS_{B}}$  

```{r}
A<-c("A1","A1","A1","A1","A2","A2","A2","A2")
B<-c("B1","B2","B1","B2","B1","B2","B1","B2")
x<-c(2,4,4,6,8,10,12,14)
xx<-as.data.frame(cbind(A,B,x))
xx$A<-as.factor(xx$A)
xx$B<-as.factor(xx$B)
xx$x<-as.numeric(xx$x)
knitr::kable(xx)
```

```{r quiz2, ECHO=F}
learnr::quiz(
 learnr::question("What is SS Between? ",
    learnr::answer("106", correct = TRUE),
    learnr::answer("98"),
    learnr::answer("8"),
    learnr::answer("20"),  
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  ),
   learnr::question("What is SS Within?",
    learnr::answer("106"),
    learnr::answer("98"),
    learnr::answer("8"),
    learnr::answer("20", correct = TRUE),  
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
),
   learnr::question("What is SS A?",
    learnr::answer("106"),
    learnr::answer("98", correct = TRUE),
    learnr::answer("8"),
    learnr::answer("20"),      
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  ),
   learnr::question("What is SS B?",
    learnr::answer("106"),
    learnr::answer("98"),
    learnr::answer("8", correct = TRUE),
    learnr::answer("20"),      
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  ),
learnr::question("What is SS A*B?",
    learnr::answer("0", correct = TRUE),
    learnr::answer("98"),
    learnr::answer("8"),
    learnr::answer("20"),      
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  ),
learnr::question("What is F A?",
    learnr::answer("19.6", correct = TRUE),
    learnr::answer("1.6"),
    learnr::answer("0"),
    learnr::answer("20"),      
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  ),
learnr::question("What is F B?",
    learnr::answer("19.6"),
    learnr::answer("1.6", correct = TRUE),
    learnr::answer("0"),
    learnr::answer("20"),      
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  ),
learnr::question("What is F AxB?",
    learnr::answer("19.6"),
    learnr::answer("1.6"),
    learnr::answer("0", correct = TRUE),
    learnr::answer("20"),      
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  ),
learnr::question("What is partial eta squared for B?",
    learnr::answer(".83"),
    learnr::answer(".00"),
    learnr::answer(".29", correct = TRUE),
    learnr::answer(".06"),      
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  )
)

``` 


## Video III Interpreting Interactions

![](https://youtu.be/GzU0Y15ZieE)

## Quiz 

```{r }
A<-c("A1","A1","A2","A2")
B<-c("B1","B2","B1","B2")
x<-c(2,2,2,4)
xx<-as.data.frame(cbind(A,B,x))
xx$A<-as.factor(xx$A)
xx$B<-as.factor(xx$B)
xx$x<-as.numeric(xx$x)
library(ggplot2)
bars <- ggplot(xx, aes(x=A, y=x, fill=B))
bars + stat_summary(fun = mean, geom="bar", position="dodge")+ggtitle("Graph1")+coord_fixed(ratio = 0.5)

x<-c(2,4,4,8)
xx<-as.data.frame(cbind(A,B,x))
xx$A<-as.factor(xx$A)
xx$B<-as.factor(xx$B)
xx$x<-as.numeric(xx$x)
bars <- ggplot(xx, aes(x=A, y=x, fill=B))
bars + stat_summary(fun = mean, geom="bar", position="dodge")+ggtitle("Graph2")+coord_fixed(ratio = 0.5)

x<-c(10,10,8,4)
xx<-as.data.frame(cbind(A,B,x))
xx$A<-as.factor(xx$A)
xx$B<-as.factor(xx$B)
xx$x<-as.numeric(xx$x)

bars <- ggplot(xx, aes(x=A, y=x, fill=B))
bars + stat_summary(fun = mean, geom="bar", position="dodge")+ggtitle("Graph3")+coord_fixed(ratio = 0.5)
```

```{r quiz3, ECHO=F}
learnr::quiz(
 learnr::question("For graph 1 which main effects appear to be produced entirely by the interaction?",
    learnr::answer("Main Effect of A"),
    learnr::answer("Main Effect of B"),
    learnr::answer("Both main effects",correct = TRUE),
    learnr::answer("Neither main effects"),  
    correct = "Correct! If the interaction was not present neither main effect would exist." ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  ),
 learnr::question("For graph 2 which main effects appear to be produced entirely by the interaction?",
    learnr::answer("Main Effect of A"),
    learnr::answer("Main Effect of B"),
    learnr::answer("Both main effects"),
    learnr::answer("Neither main effect",correct = TRUE),  
    correct = "Correct! A2 is consistently larger than A1 and B2 is consistently larger than B1",
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
),
 learnr::question("For graph 3 which main effects appear to be produced entirely by the interaction?",
    learnr::answer("Main Effect of A"),
    learnr::answer("Main Effect of B",correct = TRUE),
    learnr::answer("Both main effects"),
    learnr::answer("Neither main effecct"),  
    correct = "Correct! A2 is always lower than A1 but B2 is only lower than B1 because of the interaction" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  )
)

``` 

## Video IV: Example and R 

![](https://youtu.be/7tVJzzl-UC8)



## Factorial ANOVA in R

```{r echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Factorial ANOVA in R looks a lot like one factor ANOVA, but there is some extra code to get the correct sums of squares. There are numerous types of SS and ways to conduct ANOVA but only one type that is relevant to situations when we test interactions. 

The first line of code sets the correct contrast. 

The second conducts the ANOVA, writing it to an object (anova2F).

The third line uses the `car` package to produce the correct sums of squares. 

The final line produces the effect size using `lsr`.

```{r}
options(contrasts=c("contr.helmert", "contr.poly"))
anova2F<-lm(XPosEmo~goal_manip*fr_manip, data=Green1)
car::Anova(anova2F, type = 3)
lsr::etaSquared(anova2F)
```

### Exercise 

Using the *Green1* data, predict attitudes (**XMeta**) from **goal_manip** and **fr_manip**.

```{r ex2, exercise = TRUE, exercise.lines = 4}

```

```{r ex2-solution}
options(contrasts=c("contr.helmert", "contr.poly"))
ex<-lm(XMeta~goal_manip*fr_manip, data=Green1)
car::Anova(ex, type = 3)
lsr::etaSquared(ex)
```


### Descriptive Statistics

One last useful piece are descriptive statistics. As before the tapply command can deliver those 

```{r}
tapply(Green1$XPosEmo,list(Green1$goal_manip,Green1$fr_manip),mean)
tapply(Green1$XPosEmo,list(Green1$goal_manip,Green1$fr_manip),sd)
```

Adapt the code above to produce descriptive statistics for your analyses.

```{r ex5, exercise = TRUE, exercise.lines = 2}

```

```{r ex5-solution}
tapply(Green1$XMeta,list(Green1$goal_manip,Green1$fr_manip),mean)
tapply(Green1$XMeta,list(Green1$goal_manip,Green1$fr_manip),sd)
```

(We will graph the relationship in the next section)

## Quiz

```{r echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
options(contrasts=c("contr.helmert", "contr.poly"))
ex<-lm(XMeta~goal_manip*fr_manip, data=Green1)
car::Anova(ex, type = 3)
lsr::etaSquared(ex)
tapply(Green1$XMeta,list(Green1$goal_manip,Green1$fr_manip),mean)
tapply(Green1$XMeta,list(Green1$fr_manip),mean)
tapply(Green1$XMeta,list(Green1$goal_manip),mean)
```




```{r quiz5, ECHO=F}
learnr::quiz(
 learnr::question("What variables produced statisitcally significant effects?",
    learnr::answer("Both main effects"),
    learnr::answer("The interaction"),
    learnr::answer("Goal Manipulation and the Interaction"),
    learnr::answer("Friendship Manipulation and the Interaction",correct=TRUE),  
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  ),
 learnr::question("Which of the following best describes the main effect of the friendship manipulation?",
    learnr::answer("There was not a difference between homogenous and diverse conditions"),
    learnr::answer("The homogenous friends conditions resulted in more positive attitudes than the diverse friends condition"),
    learnr::answer("The diverse friends conditions resulted in more positive attitudes than the homogenous friends condition",correct=TRUE),
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
),
 learnr::question("Which of the following best describes the main effect of goal manipulation on attitudes?",
    learnr::answer("There was not difference between learning vs. performance goals",correct=TRUE),
    learnr::answer("Learning goals produced more favorable attitudes than did performance goals"),
    learnr::answer("Performance goals produced more favorable attitudes than did learning goals"),
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  ),
 learnr::question("Which of the following best describes the interaction effect?",
    learnr::answer("There was not an interaction"),
    learnr::answer("Diverse condition produce more favorable attitudes than did homogenous condition"),
    learnr::answer("Performance goals produce more favorable attitudes than did learning goals"),
    learnr::answer("The diverse condition produced more favorable attitudes than homogenous but only in the performance goals condition",correct=TRUE),
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  )
)

``` 


## Video 5: Simple effects

![](https://youtu.be/n4GHMzGoNSQ)

## Graphing with `ggplot2`
```{r echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Before we interpret our effects, it is useful to view them graphically. This involves some minor changes to the `ggplot2` code we used for the one factor. I highlight the new pieces of code below. 

As before group_by and summarise break down our means, standard deviations, and standard errors. For this graph, we are now doing this by cell. There are a couple of changes to accomplish this. First, the group_by command now includes both factors, separated by a comma. 

The other changes is to the ggplot code. x is = one factor, y = outcome variable, fill = the other factor. Generally, fill should be the factor with the fewest levels (not an issue here as they both have two). 

```{r}
yy<-Green1 %>%
  group_by(fr_manip,goal_manip) %>%
  summarise(mean.cell = mean(XPosEmo, na.rm = TRUE),
            sd.cell = sd(XPosEmo, na.rm = TRUE),
            n.cell = n()) %>%
            mutate(se.cell = sd.cell / sqrt(n.cell),
            lower.ci.cell = mean.cell - qt(1 - (0.05 / 2), n.cell - 1) * se.cell,
            upper.ci.cell = mean.cell + qt(1 - (0.05 / 2), n.cell - 1) * se.cell)
ggplot(yy, aes(x=fr_manip, y=mean.cell,fill=goal_manip)) + 
    geom_bar(position=position_dodge(), stat="identity") +
    geom_errorbar(aes(ymin=lower.ci.cell, ymax=upper.ci.cell),
                  width=.2,       
position=position_dodge(.9))
```

### Exercise

Adapt the `ggplot2` code to make a graph for your analysis. 

```{r ex3, exercise = TRUE, exercise.lines = 13}

```

```{r ex3-solution}
yy<-Green1 %>%
  group_by(fr_manip,goal_manip) %>%
  summarise(mean.cell = mean(XMeta, na.rm = TRUE),
            sd.cell = sd(XMeta, na.rm = TRUE),
            n.cell = n()) %>%
            mutate(se.cell = sd.cell / sqrt(n.cell),
            lower.ci.cell = mean.cell - qt(1 - (0.05 / 2), n.cell - 1) * se.cell,
            upper.ci.cell = mean.cell + qt(1 - (0.05 / 2), n.cell - 1) * se.cell)
ggplot(yy, aes(x=fr_manip, y=mean.cell,fill=goal_manip)) + 
    geom_bar(position=position_dodge(), stat="identity") +
    geom_errorbar(aes(ymin=lower.ci.cell, ymax=upper.ci.cell),
                  width=.2,       
position=position_dodge(.9))
```

## Simple effects in R

Simple effects tests are necessary to decompose interactions. With simple effects we are focusing on the effect of one factor at each individual level of the other. The `phia` package conducts these tests simply. The code for the **testinteractions** command requires an object (xx) that captures the analysis, fixed refers to the factor that you examine condition by condition, and the across command sets the variable to compare across each level of the fixed factor. 

Note: If you have a design that has more than two levels (e.g. a 2x3 ANOVA), it is usually best to set the factor with the most levels as fixed. 

```{r}
options(contrasts=c("contr.helmert", "contr.poly"))
xx<-lm(XPosEmo~goal_manip*fr_manip, data=Green1)
phia::testInteractions(xx,fixed = c("goal_manip"), across="fr_manip",adjustment="none")
```



## Video 6: Why I hate partial eta-squared

![](https://youtu.be/dns68Ls5u1A)

## Quiz 

```{r echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r quiz6, ECHO=F}
learnr::quiz(
 learnr::question("How would you calculate eta-squared?",
    learnr::answer("SS effect over SS within"),
    learnr::answer("SS effect over SS between"),
    learnr::answer("SS effect over SS total minus SS from the other effects"),
    learnr::answer("SS effect over SS total",correct=TRUE),  
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  ),
 learnr::question("Partial eta squared can be meaningfully compared across different studies",
    learnr::answer("True"),
    learnr::answer("False",correct=TRUE),    
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  ),
 learnr::question("Eta squared can be meaningfully compared across different studies",
    learnr::answer("True",correct=TRUE),
    learnr::answer("False"),      correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  )
)

```

## Video VIII: APA Style

![](https://youtu.be/ePNTkH2bYcs)

## APA Style Write-up

In this section we will review how to present findings in APA style. 

```{r}
options(contrasts=c("contr.helmert", "contr.poly"))
xx<-lm(XPosEmo~goal_manip*fr_manip, data=Green1)
car::Anova(xx, type = 3)
lsr::etaSquared(xx)
tapply(Green1$XPosEmo,list(Green1$goal_manip,Green1$fr_manip),mean)
tapply(Green1$XPosEmo,list(Green1$goal_manip,Green1$fr_manip),sd)
tapply(Green1$XPosEmo,Green1$goal_manip,mean)
tapply(Green1$XPosEmo,Green1$goal_manip,sd)
tapply(Green1$XPosEmo,Green1$fr_manip,mean)
tapply(Green1$XPosEmo,Green1$fr_manip,sd)
phia::testInteractions(xx,fixed = c("goal_manip"), across="fr_manip",adjustment="none")
```

A 2x2 between subjects ANOVA examined the relationship between goals and friendship diversity on Positive emotions. Those participants who believed they were interacting with a partner who had a diverse group of friends (*M* = 5.56, *S* = 1.30) reported more positive emotions than those interacting with a partner with homogeneous friends (*M* = 5.08, *S* = 1.22), *F*(1,144)=5.84, *p*=.02, partial $\eta^2$ = .037. There was no difference between participants given learning (*M* = 5.42, *S* = 1.26) rather than performance goals (*M* = 5.20, *S* = 1.29), *F*(1,144)=0.90, *p*=.34, partial $\eta^2$ = .007. 

These findings are qualified by a significant interaction, *F*(1,144)=7.09, *p*=.008, partial $\eta^2$ = .014. Simple effects tests reveal that in the performance conditions, interactions with a partner with diverse friends (*M* = 5.74, *S* = 1.21) related to more positive emotions than interacting with someone with homogeneous fiends (*M* = 4.71, *S* = 1.18), *F*(1,144)=12.53, *p*<.001, partial $\eta^2$ = .08. In the learning condition, interactions with a partner with diverse friends (*M* = 5.40, *S* = 1.38) produced similar positive emotions compared to interacting with someone with homogeneous fiends (*M* = 5.44, *S* = 1.15), *F*(1,144)=0.03, *p*=.86, partial $\eta^2$ = .00.

## Interaction Interpretations

Interpreting interactions can be very tricky. With that in mind, I've provided some additional practice for graphical interpretation. I have tried to provide examples that show the most common patterns you will encounter. 

```{r}
Type<-c("Control","Control", "Treatment","Treatment")
Length<-c("Short","Long", "Short","Long")
Symptoms<-c(6,3,3,6)
ggplot(yy, aes(x=Type, y=Symptoms,fill=Length)) + 
    geom_bar(position=position_dodge(), stat="identity") + labs(title="Graph 1")

Type<-c("Control","Control", "Treatment","Treatment")
Length<-c("Short","Long", "Short","Long")
Symptoms<-c(6,6,6,3)
ggplot(yy, aes(x=Type, y=Symptoms,fill=Length)) + 
    geom_bar(position=position_dodge(), stat="identity") + labs(title="Graph 2")

Type<-c("Control","Control", "Treatment","Treatment")
Length<-c("Short","Long", "Short","Long")
Symptoms<-c(6,6,3,1)
ggplot(yy, aes(x=Type, y=Symptoms,fill=Length)) + 
    geom_bar(position=position_dodge(), stat="identity") + labs(title="Graph 3")
```

### Quiz

```{r quiz7, ECHO=F}
learnr::quiz(
 learnr::question("Which interpretation best fits the first graph?",
    learnr::answer("Those in the treatment condition improved but only when provided long-term"),
    learnr::answer("Both treatment lengths improved outcomes for the treatment group, however, long-term proved more effective than short term"),
    learnr::answer("Length of treatment has differential effects for control and treatment conditions. The control group participants got better as treatment (wait time) got longer but those who received treatment did far better in the short term",correct=TRUE),
    learnr::answer("None of the above"),  
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  ),
 learnr::question("Which interpretation best fits the second graph?",
    learnr::answer("Those in the treatment condition improved but only when provided long-term",correct=TRUE),
    learnr::answer("Both treatment lengths improved outcomes for the treatment group, however, long-term proved more effective than short term"),
    learnr::answer("Length of treatment has differential effects for control and treatment conditions. The control group participants got better as treatment (wait time) got longer but those who received treatment did far better in the short term"),
    learnr::answer("None of the above"),  
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  ),
 learnr::question("Which interpretation best fits the third graph?",
    learnr::answer("Those in the treatment condition improved but only when provided long-term"),
    learnr::answer("Both treatment lengths improved outcomes for the treatment group, however, long-term proved more effective than short term",correct=TRUE),
    learnr::answer("Length of treatment has differential effects for control and treatment conditions. The control group participants got better as treatment (wait time) got longer but those who received treatment did far better in the short term"),
    learnr::answer("None of the above"),  
    correct = "Correct!" ,
    incorrect = "Sorry, that is incorrect. Try again.",
    random_answer_order = TRUE,
    allow_retry = T
  )
)

```

##
Congratulations! You've reached the end of the tutorial. 

